---
title: "BDA Project: Hurricane forecasting in Stan"
author: | 
  | José Miguel Ramírez & Jonas Lindblad
  | Aalto University
output: 
    beamer_presentation:
        incremental: true
        slide_level: 2
        keep_tex: false
        toc: false
        theme: "Singapore"
        colortheme: "beaver"
        fonttheme: "structurebold"
header-includes:
  \usepackage{dcolumn}
---

<!-- 1. introduce the problem we are trying to solve -->
# Hurricane introduction

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = '..')
```


```{r, include=FALSE}
library(tufte)
source("init.r", local = knitr::knit_global())

# images
hurricane <- file.path(getwd(), img_path, "hurricane.jpg")
path <- file.path(getwd(), img_path, "corrplot_delta_small_presentation_centered.png")

# diagrams
regression_diag <- file.path(getwd(), img_path, "simple_regression.png")
skew_diag <- file.path(getwd(), img_path, "skew_regression.png")
variance_diag <- file.path(getwd(), img_path, "variance_regression.png")

# ships variables
ships_no <- file.path(getwd(), img_path, "ships_img.png")
ships_time <- file.path(getwd(), img_path, "ships_img_time.png")
ships_vars <- file.path(getwd(), img_path, "ships_img_vars.png")

# pareto-k plots
pareto_linear_path <- file.path(getwd(), img_path, "pareto_linear.png")
pareto_skew_path <- file.path(getwd(), img_path, "pareto_skew.png")
pareto_variance_path <- file.path(getwd(), img_path, "pareto_variance.png")

# forecast plots 
dorian <- file.path(getwd(), img_path, "eval_dorian2.png")
chris <- file.path(getwd(), img_path, "eval_chris2.png")
florence <- file.path(getwd(), img_path, "eval_florence2.png")
gabrielle <- file.path(getwd(), img_path, "eval_gabrielle2.png")
helene <- file.path(getwd(), img_path, "eval_helene2.png")
isaac <- file.path(getwd(), img_path, "eval_isaac2.png")
jose <- file.path(getwd(), img_path, "eval_jose2.png")
```
<!-- note that for syntax reasons there needs to be a space '- ' -->
- Destructive storms occuring in the late summer and fall in the northern hemisphere’s tropical region.   
- Classified by their wind intensity at the eye wall.  
- They can cause extreme levels of flooding and destroy many buildings.  
- Monetary damages and loss of lives increase with an almost exponential character as a function of storm intensity.  

\center

![](`r hurricane`){width=60%}


## Hurricane forecasting basics

- Forecasters predict two quantities: track and intensity
- Two kinds of models: _dynamical_ and _statistical_
    - Dynamical: simulate the laws of physics
    - Statistical: estimate based on historical data
- Dynamical vs. statistical: good at long- and short-range forecasts respectively
- ...but for hurricane forecasts short-range is usually more interesting
- _Rapid intensification_: forecasted better by dynamical models
- This project: a _statistical_ model for _intensity_

<!-- 1.1. needed a longer explanation for this part... -->
## Hurricane forecasting basics: the SHIPS data

The US government forecasting agency, the National Hurricane Center (NHC), uses a large number of models operationally. The models (together: the _model ensemble_) are used together with experienced meteorologists' judgment to provide the official forecast. 

- Surprisingly, the best _single_ (short-range) model is a multiple linear regression!
- The NHC regression model: Statistical Hurricane Intensity Prediction Scheme (SHIPS)
- SHIPS: ~140 covariates, many calculated from data sources more easily available to the NHC
- SHIPS dataset: publically available with no restrictions (link: [SHIPS Development](http://rammb.cira.colostate.edu/research/tropical_cyclones/ships/index.asp))
- ...but the documentation is terrible
- SHIPS: only a point estimate; our project: a predictive distribution

<!-- 1.2-a. an image explaining the variable set we have used -->
## Hurricane forecasting basics: the SHIPS data

The SHIPS developmental data is confusing!

![](`r ships_no`){width=100%}

## Hurricane forecasting basics: the SHIPS data

We are making _synoptic_ models.

![](`r ships_time`){width=100%}

## Hurricane forecasting basics: the SHIPS data

We are making _synoptic_ models and choosing variables.

![](`r ships_vars`){width=100%}

<!-- 1.2-b. correlation plot ? -->
## Hurricane forecasting basics: our selection

We have not done statistical variable selection. Choice of variable subset is based on theory.

>- **CSST**: (climatological) sea surface temperature
>- **RHLO**: low-altitude relative humidity
>- **T200**: air temperature at 200 mb (very high altitude)
>- **SHRD**: wind shear between 850 and 200 mb
>- **VMPI**: maximum potential intensity

:::: {.columns}
::: {.column width="40%"}
![](`r path`){width=90%}
:::
::: {.column width="60%"}
>- for testing, we have variable sets *A, B, C*
>- *A*: LAT/LON, VMAX, CSST, SHRD
>- *B*: LAT/LON, VMAX, CSST, SHRD, VMPI
>- *C*: LAT/LON, VMAX, CSST, SHRD, VMPI, RHLO, T200
:::
::::

# Intensity change predictive model

<!-- 2. explain the 3 models; visually rather than with text -->
<!-- idea: Kruschke diagrams -->
<!-- https://github.com/rasmusab/distribution_diagrams -->
<!-- we would need 3 different Kruschke diagrams for the models -->


The SHIPS Blunder: a simple linear regression

![](`r regression_diag`){width=95%}

## Intensity change predictive model

Model 2: regression with skewness

- errors not symmetric around the mean prediction!
- rapid intensification!

![](`r skew_diag`){width=75%}

## Intensity change predictive model

Model 3: regression with a linear model for standard deviation

- fewer storms reach higher values of `VMAX` 
- allow for higher variance to account for larger historical uncertainty

![](`r variance_diag`){width=75%}

<!-- 2.1. some remarks on limits of the model(s) -->
## Models: some remarks

- SHIPS: predict $V_{max}$; our models: predict $\Delta V_{max}$
- we also decided to include latitude and longitude as variables; geographic region is known to be important for storm development
- reminder: we do 12-hour forecasts
- we standardized all of our data; priors chosen to be weak in standardized scale
- SHIPS data: 1982-2019; our restriction: 2017-2019
- poor problem setup! True model is the laws of physics, but we are fitting a regression
- models were programmed in Stan; sampling with `rstan` resulted in no divergences or issues except for the skew model and the issue was solved by increasing max tree depth to 15

<!-- 3.1.1 compare with other models -->
# Forecasting: Model Comparison

\vspace{0.4cm}

Dataset comparison for the **linear regression model** (LOOCV)  

\vspace{0.1cm}

\begin{table}[!h]
\fontsize{8pt}{12pt}\selectfont
\begin{tabular}{ c c c }
 Dataset & \textbf{elpd\_diff} & \textbf{se\_diff} \\ \hline
 \textbf{\textit{C}} & 0.0 & 0.0 \\  
 \textit{B} & -25.0 & 6.5 \\  
 \textit{A} & -27.4 & 6.3 \\     
\end{tabular}
\end{table}

\vspace{-0.2cm}

\center
![](`r pareto_linear_path`){height=60%}

<!-- 3.1.2 compare with other models -->
## Forecasting: Model Comparison (2)

\vspace{0.4cm}

Dataset comparison for the **skewed regression model** (LOOCV)  

\vspace{0.1cm}

\begin{table}[!h]
\fontsize{8pt}{12pt}\selectfont
\begin{tabular}{ c c c }
 Dataset & \textbf{elpd\_diff} & \textbf{se\_diff} \\ \hline
 \textbf{\textit{C}} & 0.0 & 0.0 \\  
 \textit{B} & -23.2 & 6.2 \\  
 \textit{A} & -28.7 & 6.2 \\     
\end{tabular}
\end{table}


\vspace{-0.2cm}
\center
![](`r pareto_skew_path`){height=60%}

<!-- 3.1.3 compare with other models -->
## Forecasting: Model Comparison (3)

\vspace{0.4cm}

Dataset comparison for the **Changing variance model** (LOOCV)  

\vspace{0.1cm}  

\begin{table}[!h]
\fontsize{8pt}{12pt}\selectfont
\begin{tabular}{ c c c }
 Dataset & \textbf{elpd\_diff} & \textbf{se\_diff} \\ \hline
 \textbf{\textit{C}} & 0.0 & 0.0 \\  
 \textit{B} & -32.6 & 8.2 \\  
 \textit{A} & -37.1 & 8.2 \\     
\end{tabular}
\end{table}


\vspace{-0.2cm}
\center
![](`r pareto_variance_path`){height=60%}

<!-- 3.1.4 compare with other models -->
## Forecasting: Model Comparison (4)

Model comparison using the Dataset C (LOOCV) 
\vspace{1cm}

\begin{table}[!h]
\renewcommand{\arraystretch}{1.25}
\begin{tabular}{ c c c }
 Model & \textbf{elpd\_diff} & \textbf{se\_diff} \\ \hline
 \textbf{Variance} & 0.0 & 0.0 \\  
 Skew & -176.3 & 27.8 \\  
 Linear & -205.5 & 34.9 \\     
\end{tabular}
\end{table}

<!-- Marginal plot for the theta-variables -->
## Marginal posteriors

>- $\theta_1$: constant term
>- $\theta_2, \theta_3$: latitude, longitude
>- $\theta_4$: sea surface temperature (CSST)
>- $\theta_5$: relative humidity (RHLO)
>- $\theta_6$: wind shear (SHRD)
>- $\theta_7$: maximum potential intensity (VMPI)
>- $\theta_8$: air temperature at 200 mb (T200)
>- $\theta_9$: intensity at storm core (VMAX)

<!-- 3.2. try looking for historical NHC predictions? (can be skimmed incase of time running out) -->
## Forecasting: checking predictions

Hurricane Dorian 2019. The image shows a 90% credible interval. Black dotted line: true VMAX
\center

![](`r dorian`){height=70%}

## Forecasting: checking predictions

Hurricane Florence 2018. The image shows a 90% credible interval. Black dotted line: true VMAX
\center

![](`r florence`){height=70%}

## Forecasting: checking predictions

Hurricane Helene 2018. The image shows a 90% credible interval. Black dotted line: true VMAX
\center

![](`r helene`){height=70%}



<!-- 4. Final section -->
# Concluding section

Further development ideas:

- variable selection in full SHIPS dataset
- more time series autoregressive components
- use LGEM model (will explain)


<!-- 4.1. Conclusions  -->
## A final takeaway
<!-- don't put your email address directly : github crawlers will find it and send you spam -->

> "The hurricane will be moving through an environment of low vertical
wind shear, a moist mid-level atmosphere, and increasing upper-ocean
heat content, and this is ideal for some additional intensification.
However, given that the SHIPS models do not show any significant
change in the intensity, the NHC forecast keeps Irma a powerful
hurricane through five days."
> `r tufte::quote_footer('--- NHC 5AM advisory Sep. 5th')`

While the NHC published the advisory, Irma was undergoing rapid intensification from 130 to 175 knots over a span of only a few hours.

## Conclusions & contact info

>- Takeaway: the SHIPS model is terrible!
>- Simple changes to the predictive distribution can improve the model
>- Further development using Bayesian methods seems promising and there are several possible directions

More at our Github repo: [https://github.com/jnlb/bda-hurricane-modeling](https://github.com/jnlb/bda-hurricane-modeling)

The SHIPS website: [http://rammb.cira.colostate.edu/research/tropical_cyclones/ships/index.asp](http://rammb.cira.colostate.edu/research/tropical_cyclones/ships/index.asp)

Contact info: 

:::: {.columns}
::: {.column width="50%"}
José Miguel Ramírez
:::
::: {.column width="50%"}
Jonas Lindblad

>- rocket-chat: `@jnlb`
:::
::::

<!-- 4.2 Extra material -->
## Additional information
The SHIPS model:
$$ y_{i} \sim \mathcal{N}(\alpha + X_i \cdot\beta_{N-1}, \sigma), \ i=1,\dots,r, $$
where we let $X_i$ denote the $i$:th row of the data, $\beta_{N-1}$ is an $N-1$-dimensional parameter vector, and $r$ is the number of observations (rows) in the data. Its priors were 
$$\begin{bmatrix} \alpha_0 \\ \beta_{N-1,0} \end{bmatrix} \sim \mathcal{N}(\mathbf{0}_N, 10 \cdot \mathbf{I}_N), \sigma_0 \sim \textrm{Inv-}\chi^2(\tfrac{1}{10}) .$$
The skew-normal regression model: 
$$ y_{i} \sim \textrm{SkewNormal}(\alpha + X_i \cdot\beta_{N-1}, \sigma, \psi), \ \ i=1,\dots,r, $$
with priors 
$$\begin{bmatrix} \alpha_0 \\ \beta_{N-1,0} \end{bmatrix} \sim \mathcal{N}(\mathbf{0}_N, 10 \cdot \mathbf{I}_N), \sigma_0 \sim \textrm{Inv-}\chi^2(\tfrac{1}{10}) , \ \psi_0 \sim \mathcal{N}(0,1) .$$

## Additional information

The variance model:
$$ y_{i} \sim \mathcal{N}(\alpha + X_i \cdot\beta_{N-1}, \sigma + \gamma\vert V_{max,i} \vert ), \ i=1,\dots,r, $$
where, again, we use the same notation as before and let $V_{max,i}$ denote the $V_{max}$-value of the $i$:th row. The priors that were fed into Stan were
$$\begin{bmatrix} \alpha_0 \\ \beta_{N-1,0} \end{bmatrix} \sim \mathcal{N}(\mathbf{0}_N, 10 \cdot \mathbf{I}_N), \ \sigma_0 \sim \textrm{Inv-}\chi^2(\tfrac{1}{10}), \ \gamma_0 \sim \Gamma(1,1) .$$

## Additional information

The variance regression was run in `rstan` with the following options
```
variance_m <- rstan::stan_model(file = file.path(mod_path, 
              "minimal3.stan"))
variance_model <- rstan::sampling(variance_m, 
                data = stan_data, 
                control = list(max_treedepth = 10),
                iter=4000, seed = SEED)
```

<!-- We need to consider what questions the TAs are likely to ask -->

<!-- Consider including an image which explains how Hurricanes form -->
