---
title: "BDA Project: Hurricane forecasting in Stan"
author: "José Miguel Ramírez & Jonas Lindblad"
output: 
    beamer_presentation:
        incremental: true
---

<!-- 1. introduce the problem we are trying to solve -->
## Hurricane forecasting basics

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

- Forecasters predict two quantities: track and intensity
- Two kinds of models: _dynamical_ and _statistical_
    - Dynamical: simulate the laws of physics
    - Statistical: estimate based on historical data
- Dynamical vs. statistical: good at long- and short-range forecasts respectively
- _Rapid intensification_: forecasted better by dynamical models
- This project: a _statistical_ model for _intensity_

<!-- 1.1. an image explaining the variable set we have used -->
## Hurricane forecasting basics: the SHIPS data


<!-- 1.2. correlation plot ? -->
## Hurricane forecasting basics: our selection


<!-- 2. explain the 3 models; visually rather than with text -->
## Intensity change predictive model


<!-- 2.1. some remarks on limits of the model(s) -->
## Model: limitations


<!-- 2.2. model comparison with posterior predictive checks (visually) -->
## Model: posterior predictive checking


<!-- 2.3. data selection comparisons -->
## Model: variables to use


<!-- 2.4. distribution of coefficients (visually) -->
## Model: marginal posteriors of coefficients


<!-- 2.5. special case: bivariate margin of SHRD and CSST? (optional?) -->
## Model: margin of SHRD and CSST


<!-- 3. example forecasts on historical storms -->
## Forecasting


<!-- 3.1. compare with other models -->
## Forecasting: model comparison


<!-- 3.2. try looking for historical NHC predictions? (can be skimmed incase of time running out) -->
## Forecasting: what about the NHC?


<!-- 4. Final section -->
## Problems to solve & development ideas


<!-- 4.1. Conclusions  -->
## Conclusions & contact info
<!-- don't put your email address directly : github crawlers will find it and send you spam -->


<!-- 4.2 Extra material -->
## Additional information

<!-- We need to consider what questions the TAs are likely to ask -->

<!-- Consider including an image which explains how Hurricanes form -->


<!-- EXTRA: beamer syntax tutorial below -->
## R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## Slide with R Output

```{r cars, echo = TRUE}
summary(cars)
```

## Slide with Plot

```{r pressure}
plot(pressure)
```

